Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure

### 2015-09-03 3dpas_intro.tex

\address{\affilnum{1}Rutgers University\break
\affilnum{2}University of Edinburgh\break
\affilnum{3}University of St.~Andrews\break
\affilnum{4}University of Chicago\break
\affilnum{5}Argonne National Laboratory\break
\affilnum{6}Cardiff University\break
\affilnum{7}University of Southern California\break
}

@@

Pan-STARRS, a contemporary sky survey with similar goals as LSST,  uses a scripted programming model
on a cluster for image processing of telescope frames to extract object attributes as a CSV
file.
%\alnote{Is the workflow part still part of the app as defined before with the stages: data transformation \& decision and storage?} \katznote{I think this is data transformation.}\alnote{ok}
This is followed by followed by scientific workflows, based on Trident, that load the files
into individual databases daily, and merges them with a master distributed-database on a weekly basis on an
HPC cluster. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
language with user defined functions and submitted for execution on the master database using a batch system.

@@

The application~\cite{modisAzure} is used to handle the processes of downloading data from different
sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
spatial and temporal resampling, and gap filling. The application itself is a loosely-coupled data
pipeline that coordinates these processes.

@@

{\em Data Sources} are used in the acquisition stage as inputs of data
into the application. Broadly speaking they include types such as
files, servers, sensors, instruments, simulations and models.
Additionally, databases are often considered as an intermediate data
source in types of system. The dynamic characteristics are primarily
due to availability. These are often accessed in the Data Generation stage of the application
(\figurename~\ref{fig:figures_application-stages}).

{\em Data Storage} components are used to provide persistent access to
data by the application. They may be structured (e.g., databases,
No-SQL tables) or unstructured (e.g., collections of audio
recordings). The dynamic characteristics are primarily due to the
level of persistence, e.g., temporary, durable, permanent (archived or
published), or replicated; and also in the structure of the data
stored, e.g., schema free tables as opposed to relational tables. There
may also be dynamic characteristics of storage due to rate of growth
of data, e.g., a field survey may bring in a lot of data during summer
but none during other times. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
but may also be present at the boundaries between the stages for transient staging of data.

@@

Revisiting the overall issue of infrastructure, which we defined as
the hardware and software needed to run the applications, we
note that the specific hardware isn't a primary issue.  The
applications we've discussed all use a mix of the systems that exist
today, which have aspects of web, grid, and cloud, and can be
considered a hybrid of the three.  The key infrastructure components
are really the software and services on top of the hardware.  In this
section, we've discussed infrastructure to support data management and
data analysis, as well as the tools and systems used to program
applications, and the specific components used for the application in
\S\ref{sec:scenarios}.  However, there is not much commonality in the
software and services that the applications use, so the software and programming infrastructure
we've described here is much more broad than what is used by the
applications we've studied.

@@

This section brings together some key observations
from the applications and infrastructure
analysis carried out in this paper.
%as take away messages. 
Section~\ref{sec:observations}
summarize the dynamic and distributed characteristics of
data associated with the applications we've studied in this work,
attempting to infer more general characteristics that %can be applied
may apply
to other similar applications.  In \S\ref{sec:discussion}, we
suggest two metrics that may
be associated with such applications, primarily as a basis to compare
how compute and data processing (in terms of their number and
distribution) may be used to find common features between them. Such
feature analysis is useful, we believe, for two main reasons: (i) to
identify similarities between the applications covered in
\S\ref{sec:scenarios} with others that are in the design phase or
under development, so that common requirements can be identified; and
(ii) to identify gaps in current software tools and programming
systems that such applications use, thereby leading to the
development of new tools and systems. As our application and
infrastructure coverage is, by necessity, limited, identifying common
features also enables the lessons learned to be generalized and
applied more widely. We subsequently discuss a set of architectures
for D3 science, and pose some questions we believe are pertinent as
the community begins considering the next-generation
architectures for D3 applications. In
\S\ref{sec:conclusion} we conclude by providing a number of
observations to inform applications and
cyberinfrastructure developers, based on common trends found in our study.

@@

\subsection{Observations}
\label{sec:observations}

In this subsection we summarize the source and nature of dynamic data that we observe in
the application surveyed in \S\ref{sec:scenarios}, and the types of
data distribution that we observe in such applications.

@@

From \S\ref{sec:distTypes} and
\figurename~\ref{fig:figures_application}, we find that applications
with distributed data differ based on the amount or volume of data
distributed, the degree of distribution (scattered over one site or
many sites), and the dynamic aspects associated with distribution,
e.g., latency and relative scheduling flexibility (and thus tolerance
or constraints on data transfer/aggregation).  Applications processing
the largest volumes of data (O(PB)) typically run on specialized data
cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
Fusion and NGS that run on smaller data sets mainly utilize HPC
environments. On the other end of the spectrum are highly dynamic
applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor
network applications. These applications process much less data;
often, sensor data is also archived and aggregated for historical
analyses, which leads to higher data volumes.

@@

Now that we have discussed dynamic data and distributed data in the
context of specific applications, we can use this experience to
understand the characteristics that can be associated with D3 science
applications, and the characteristics of a computational
infrastructure that may be most suited for D3 science
applications. 
We believe these are general
trends that can be applied to other applications we have not
considered.

>>>>

\address{\affilnum{1}Rutgers University\break
\affilnum{2}University of Edinburgh\break
\affilnum{3}University of St.~Andrews\break
\affilnum{4}University of Chicago\break
\affilnum{5}Argonne National Laboratory\break
\affilnum{6}Cardiff University\break
\affilnum{7}Indian Institute of Science\break
}

@@

Pan-STARRS, a contemporary sky survey with similar goals as LSST,  uses a scripted programming model
on a cluster for image processing of telescope frames to extract object attributes as a CSV
file.
%\alnote{Is the workflow part still part of the app as defined before with the stages: data transformation \& decision and storage?} \katznote{I think this is data transformation.}\alnote{ok}
This is followed by followed by scientific workflows, based on Trident, that load the files
into individual databases daily, and merges them with a master distributed-database on a weekly basis on an
HPC cluster~\cite{Simmhan:Building:2009}. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
language with user defined functions and submitted for execution on the master database using a batch system.

@@

The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
spatial and temporal resampling, and gap filling. The application itself is a loosely-coupled data
pipeline that coordinates these processes.

@@

{\em Data Sources} are used in the acquisition stage as inputs of data
into the application. Broadly speaking these sources include types such as
files, servers, sensors, instruments, simulations and models.
Additionally, databases are often considered as an intermediate data
source in some systems. The dynamic characteristics are primarily
due to availability. These are often accessed in the Data Generation stage of the application
(\figurename~\ref{fig:figures_application-stages}).

{\em Data Storage} components are used to provide persistent access to
data by the application. They may be structured (e.g., databases,
No-SQL tables) or unstructured (e.g., collections of audio
recordings). The dynamic characteristics are primarily due to the
level of persistence, e.g., temporary, durable, permanent (archived or
published), or replicated; and also in the structure of the data
stored, e.g., schema free tables as opposed to relational tables. There
may also be dynamic characteristics of storage due to rate of growth
of data, e.g., a field survey may bring in a lot of data during summer
but none during other times. Some storage components like files and databases may also serve as data
sources. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
but may also be present at the boundaries between the stages for transient staging of data.

@@

Revisiting the overall issue of infrastructure, which we defined as
the hardware and software needed to run the applications, we
note that the specific hardware is not a primary issue.  The
applications we have discussed all use a mix of the systems that exist
today, which have aspects of web, grid, and cloud, and can be
considered a hybrid of the three.  The key infrastructure components
are really the software and services on top of the hardware.  In this
section, we have discussed infrastructure to support data management and
data analysis, as well as the tools and systems used to program
applications, and the specific components used for the application in
\S\ref{sec:scenarios}.  However, there is not much commonality in the
software and services that the applications use, so the software and programming infrastructure
we have described here is much more broad than what is used by the
individual applications we have studied.

@@

This section brings together some key observations
from the applications and infrastructure
analysis carried out in this paper.
%as take away messages. 
Section~\ref{sec:discussion}
provides a discussion of dynamic and distributed characteristics of
data associated with the applications we analyze in this work,
attempting to infer more general characteristics that %can be applied
may apply
to other similar applications.  We also suggest two metrics that may
suggest two metrics that may
be associated with such applications, primarily as a basis to compare
how compute and data processing (in terms of their number and
distribution) may be used to find common features between them. Such
feature analysis is useful, we believe, for two main reasons: (i) to
identify similarities between the applications covered in
\S\ref{sec:scenarios} with others that are in the design phase or
under development, so that common requirements can be identified; and
(ii) to identify gaps in current software tools and programming
systems that such applications use, thereby leading to the
development of new tools and systems. As our application and
infrastructure coverage is, by necessity, limited, identifying common
features also enables the lessons learned to be generalized and
applied more widely. We subsequently discuss a set of architectures
for D3 science, and pose some questions we believe are pertinent as
the community begins considering the next-generation
architectures for D3 applications. In
\S\ref{sec:conclusion} we conclude by providing a number of
observations to inform applications and
cyberinfrastructure developers, based on common trends found in our study.

@@

In this section we attempt to summarize four key themes from previous
sections: (i) the source and nature of dynamic data that we observe in
the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
data distribution that we observe in such applications; (iii) the
general characteristics that can be associated with D3 science
applications; (iv) the characteristics of a computational
infrastructure that may be most suited for D3 science
applications. Whereas (i) and (ii) are primarily observations that we
can make on existing applications that we have discussed in
\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
trends that can be applied to other applications we have not
considered. \katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}

%\katznote{Some text is needed here to introduce the discussion subsection.}
%\rananote{Added some text above -- does this make sense Dan?}

@@

From \S\ref{sec:distTypes} and
\figurename~\ref{fig:figures_application}, we find that applications
with distributed data differ based on the amount or volume of data
distributed, the degree of distribution (scattered over one site or
many sites), and the dynamic aspects associated with distribution,
e.g., latency and relative scheduling flexibility (and thus tolerance
or constraints on data transfer/aggregation).  Applications processing
the largest volumes of data (O(PB)) typically run on specialized data
cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
Fusion and NGS that run on smaller data sets mainly utilize HPC
environments. On the other end of the spectrum are highly dynamic
applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor
network applications. These applications process much less data;
often, sensor data is also archived and aggregated for historical
analyses, which leads to higher data volumes. \katznote{not sure about the `other' part of this...}

@@
